---
title: "Arbres de décision et méthodes ensemblistes"
subtitle: "TP: introduction to the boosting"
author: "Audrey Poterie"
date:  # Dates
output: # Différents formats de sorties 
  # Si on compile en pdf
  # Si on compile en html
  pdf_document:
    toc: no # Montre-t-on la table des matières?
    number_sections: yes # Numérote t'on les sections?
    highlight: tango # Apparence du code
  html_document: 
    theme: journal # Theme (voir https://www.datadreaming.org/post/r-markdown-theme-gallery/)
    toc: yes # Montre-t-on la table des matières?
    number_sections: yes # Numérote t'on les sections?
    toc_float: yes # La table des matières "flotte à droite" (seulement en html)
    highlight: tango # Apparence du code
header-includes:
  - \renewcommand{\contentsname}{Table des matières} # Nom de la table des matieres
  - \newcommand{\bX}{\mathbf X}
  - \newcommand{\bx}{{\mathbf x}}
  # Inclusion d'options (si on compile en pdf)
encoding: UTF-8 # Encodage
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
rm(list = ls()) # On nettoie l'espace de travail
# Options pour le rmarkdown
knitr::opts_chunk$set(echo = TRUE, warning = FALSE,
                      comment = NA, 
                      eval = FALSE, message = FALSE)
```


# Libraries used {-}


```{r libraries}
library(gbm) # Functional Gradient Boosting
library(xgboost) # XGBoost
library(caret)
library(psych)
library(randomForest)
```

Use the documentation about the library \texttt{gbm}: \url{https://cran.r-project.org/web/packages/gbm/index.html}


Use the documentation about the library \texttt{xgboost}: \url{https://cran.r-project.org/web/packages/xgboost/xgboost.pdf}

To get the same results than the shared correction, use the following random number generator: 
```{r random_control}
seed<-2026
```

# Exercise {-}


We consider the \textit{spam} data with 4601 observations (in package \texttt{kernlab}, see documentation \url{https://archive.ics.uci.edu/ml/datasets/spambase}). Each observation is an email. There are 58 variables including one variable named \texttt{type} that indicates if an email is a spam or not.  
The goal here is to build a model that enables to predict if an email is a spam or not.

```{r spam_data}
library(kernlab)
data(spam)
dim(spam)
```



- **Question 1 [code + interpretation].** Load the data and have a look at it. 
  - How many spam are there ?  Have a look at the input variables ? What do they represent ? 
  - Then, split the dataset into a train set (70\%) and a test set (30\%). **Your procedure must be reproducible**.
```{r}
library(rsample) # to split the dataset into a test set and a train set
set.seed(seed)
data_split <- initial_split(
  data = spam,# The dataset to be split A COMPLETER EN RESPECTANT LA CONSIGNE DE LA QUESTION
  prop = 0.7,#Proportion dans le train, A COMPLETER EN RESPECTANT LA CONSIGNE DE LA QUESTION
  strata = type,# Stratification variable A COMPLETER EN RESPECTANT LA CONSIGNE DE LA QUESTION
)

train<- training(data_split)
test<- testing(data_split)

## Autre méthode 
#set.seed(seed)
#ind_test <- sample(1:nrow(spam), floor(0.3*nrow(spam)), prob=NULL)
#test   <- spam[ind_test, ]
#train   <- spam[-ind_test, ]
```

```{r}
table(train$type)
table(test$type)
```



- **Question 2 [interpretation].** First, we want to apply the \texttt{Adaboost} algorithm on the train set by using the package \texttt{gbm}. We want that our boosting model, called \texttt{ada1}, has:
  - 500 trees, 
  - a maximal tree depth = 2,
  - shrinkage parameter = $0.05$,
  - as we will want to calibrate the number of trees, we want that the
  function also performed 5-fold cross-validation. 
Run the following code. What is the problem? Use the documentation of the function to solve it.

```{r ada1}
train$type <- ifelse(train$type == "spam",1,0)  
test$type <- ifelse(test$type == "spam",1,0)
ada1 <- gbm(type~.,
            data=train,
            distribution="adaboost",
            interaction.depth=2,
            cv.fold=5,
            shrinkage=0.05,
            n.trees=500)
```


- **Question 2 bis [interpretation + code].** Explain the model\texttt{ada1}. 
  minimise le risque empirique
  - Give the problem for which the model is the solution. 
  classification binaire
  - Why do we need to use a random number generator ?
  reproductatibilité
  - Look at the output of the model and interpret it. 
  37 des 57 prédicteurs qui influence
  
```{r}
summary(ada1)
```


- **Question 3 [interpretation].** Select the optimal number of trees by using 5-fold cross-validation with the function \texttt{gbm.perf}. 

  - What criterion is used to select the the number of trees ? (See the R
  documentation of \texttt{gbm.perf}).
 
```{r cv}
calib<-gbm.perf(ada1);calib
# Remark: set.seed(seed) is not necessary here because randomness has been already fixed when you define `ada1`.
```
	
- **Question 4 [code + interpretation].** Built a third boosting model, called \texttt{ada2}, using the optimal number of trees (for the other parameters keep the same values than those used for \texttt{ada1}). Interpret the results.

```{r}
ada2 <- gbm(type~.,
            data=train,
            distribution="adaboost",
            interaction.depth=2,
            cv.fold=5,
            shrinkage=0.05,
            n.trees=calib)
min_err<-gbm.perf(ada2);min_err
```


- **Question 5 [interpretation + code].** We will now test several values for the shrinkage parameters: 0.1, 0.5. For each value, we will build a boosting model with the same parameter than \texttt{ada1} (except for the shrinkage parameters) and we will select the best number of trees by using 5-fold cross-validation. Call these models \texttt{ada3} and \texttt{ada4}. Interpret the results.

```{r}
ada3 <- gbm(type~.,
            data=train,
            distribution="adaboost",
            interaction.depth=2,
            cv.fold=5,
            shrinkage=0.1,
            n.trees=500)
ada4 <- gbm(type~.,
            data=train,
            distribution="adaboost",
            interaction.depth=2,
            cv.fold=5,
            shrinkage=0.5,
            n.trees=500)
```

  
  
- **Question 6 [interpretation + code].** Select your final boosting model by using the results you get with the 5-fold cross-validation. Explain.

```{r}
gbm.perf(ada1)
gbm.perf(ada2)
gbm.perf(ada3)
gbm.perf(ada4)
min(c(
ada1$cv.error[gbm.perf(ada1)],
ada2$cv.error[gbm.perf(ada2)],
ada3$cv.error[gbm.perf(ada3)],
ada4$cv.error[gbm.perf(ada4)]))
```

```{r}
which.min(c(
  ada1$cv.error[gbm.perf(ada1)],
  ada2$cv.error[gbm.perf(ada2)],
  ada3$cv.error[gbm.perf(ada3)],
  ada4$cv.error[gbm.perf(ada4)]
))
```



- **Question 7 [interpretation + code].** Compute the performance of the model `ada5` on the test set. Comment the results.

```{r ada5 pred}
ada5 <- ada3
ada5_pred <- predict(
  ada5,# add the model,
  newdata=test,# add the data 
  n.trees=,# specify the optimal number of trees
  type = "response", 
  single.tree = FALSE)
ada5_pred # we get the predictied probabilities
```

```{r ada5 pred b}
# We now computed the predicted class by using the predicted probabilities
# We use the classical threshold = 0.5
ada5_prediction<-ifelse(ada5_pred<0.5,0,1);class(ada5_prediction)
ada5_prediction
caret::confusionMatrix(table(ada5_prediction,test1$type), reference=1)
```



- **Question 8 [interpretation + code].** Now, we will use the package `caret` to simultaneously calibrate the following parameters : 
  - n.trees,
  - interaction.depth,
  - shrinkage.
We will use the follwong procedure:
  - 1) we first define a grid with all the values that we want to assess
  - 2) each combination of values is assessed by repeated 5 times a 3-fold 
  crossvalidation. We choose to use the AUC as performance
  criterion.
Run the code and comment the results.
  
```{r}

# Definition of the grid
grid <- expand.grid(
  n.trees = c(100, 250, 500),         # tree number
  interaction.depth = c(1, 3, 5, 7),  # maximal tree depth
  shrinkage = c(0.01, 0.05, 0.1, 0.2),# learning rate
  n.minobsinnode=10 # default value
)

# Definition of the evalaution strategy (here the crossvalidation)
ctrl <- trainControl(
  method = "repeatedcv",
  number = 3,
  classProbs = TRUE,
  summaryFunction = twoClassSummary,
  repeats=5
)

```

```{r}
# Run the procedure
set.seed(seed)
gbm_tune <- train(
  type ~ ., 
  data = train,
  method = "gbm",
  distribution = "adaboost", # for binary classification
  trControl = ctrl,
  tuneGrid = grid,
  metric = "ROC",  # criterion : AUC
  verbose = FALSE
)

# Résultats
print(gbm_tune)
plot(gbm_tune)
```


**Output**
```{block, correction_8, type = "Correction", echo = TRUE}
Stochastic Gradient Boosting 

3220 samples
  57 predictor
   2 classes: 'nonspam', 'spam' 

No pre-processing
Resampling: Cross-Validated (3 fold, repeated 5 times) 
Summary of sample sizes: 2147, 2146, 2147, 2146, 2147, 2147, ... 
Resampling results across tuning parameters:

  shrinkage  interaction.depth  n.trees  ROC        Sens       Spec     
  0.01       1                  100      0.9293683  0.9831889  0.6055162
  0.01       1                  250      0.9532915  0.9722196  0.7713160
  0.01       1                  500      0.9671066  0.9654546  0.8318361
  0.01       3                  100      0.9513458  0.9677090  0.8012608
  0.01       3                  250      0.9698039  0.9641227  0.8628842
  0.01       3                  500      0.9777025  0.9639167  0.8865248
  0.01       5                  100      0.9578645  0.9639165  0.8242711
  0.01       5                  250      0.9739257  0.9634046  0.8702916
  0.01       5                  500      0.9798217  0.9640194  0.8958235
  0.01       7                  100      0.9614413  0.9629937  0.8379827
  0.01       7                  250      0.9759591  0.9635070  0.8784870
  0.01       7                  500      0.9808149  0.9630968  0.9018125
  0.05       1                  100      0.9670398  0.9650449  0.8307329
  0.05       1                  250      0.9763301  0.9665825  0.8786446
  0.05       1                  500      0.9782325  0.9669923  0.8948779
  0.05       3                  100      0.9771664  0.9626868  0.8858944
  0.05       3                  250      0.9806072  0.9637115  0.9030733
  0.05       3                  500      0.9822298  0.9620707  0.9112687
  0.05       5                  100      0.9791804  0.9623788  0.8923562
  0.05       5                  250      0.9821867  0.9621742  0.9117415
  0.05       5                  500      0.9832285  0.9630966  0.9163121
  0.05       7                  100      0.9804609  0.9626871  0.9003940
  0.05       7                  250      0.9829284  0.9634045  0.9115839
  0.05       7                  500      0.9836050  0.9626867  0.9188337
  0.10       1                  100      0.9746097  0.9663785  0.8690307
  0.10       1                  250      0.9776638  0.9666846  0.8940898
  0.10       1                  500      0.9790859  0.9623787  0.9010244
  0.10       3                  100      0.9797504  0.9638138  0.9019701
  0.10       3                  250      0.9821506  0.9617638  0.9118991
  0.10       3                  500      0.9825752  0.9596116  0.9167849
  0.10       5                  100      0.9814633  0.9605333  0.9084318
  0.10       5                  250      0.9827928  0.9601237  0.9153664
  0.10       5                  500      0.9828361  0.9600216  0.9215130
  0.10       7                  100      0.9819259  0.9613533  0.9122143
  0.10       7                  250      0.9832119  0.9624825  0.9188337
  0.10       7                  500      0.9827561  0.9618671  0.9262411
  0.20       1                  100      0.9769781  0.9650441  0.8899921
  0.20       1                  250      0.9787372  0.9627889  0.9019701
  0.20       1                  500      0.9785306  0.9597144  0.9037037
  0.20       3                  100      0.9806292  0.9603268  0.9084318
  0.20       3                  250      0.9818339  0.9597136  0.9155240
  0.20       3                  500      0.9814960  0.9585869  0.9197794
  0.20       5                  100      0.9818900  0.9598154  0.9082742
  0.20       5                  250      0.9821569  0.9582783  0.9185185
  0.20       5                  500      0.9817585  0.9571507  0.9234043
  0.20       7                  100      0.9819747  0.9597125  0.9136328
  0.20       7                  250      0.9822338  0.9587907  0.9249803
  0.20       7                  500      0.9816315  0.9578684  0.9249803

Tuning parameter 'n.minobsinnode' was held constant at a value of 10
ROC was used to select the optimal model using the largest value.
The final values used for the model were n.trees = 500, interaction.depth = 7, shrinkage = 0.05 and n.minobsinnode = 10.
```



- **Question 9 [interpretation + code].** Construct the boosting model with the optimal values selected in the previous question and select the optimal number of trees by using 5-fold crossvalidation and the fucntion `gbm.perf`. Compute the performance of the model on the test set. Comment the results.

```{r }
# We now computed the predicted class by using the predicted probabilities
# We use the classical threshold = 0.5
ada6_prediction<-ifelse(ada6_pred<0.5,0,1);class(ada6_prediction)
ada6_prediction
caret::confusionMatrix(table(ada6_prediction,test1$type), reference=1)
```

- **Question 10 [interpretation + code].** Compare your two calibrated boosting models. Which one do you choose? Why?


 **Question 11 [interpretation + code].** Based on the selected boosting model, plot the importance of each variable using:
  - 1) the relative.influence,
  - 2) and the permutation importance (MDA).
Run the code, then interpret and compare the results.

```{r}
par(mfrow=c(1,2))
```

```{r}
par(cex.axis = 0.55, mar = c(10, 4, 4, 2))
summary(
  ada6,
  cBars = 58,
  method = relative.influence,
  las = 1
)
```

```{r}
par(cex.axis = 0.55, mar = c(10, 4, 4, 2))
set.seed(seed)
summary(
  ada6, 
  cBars = 58,
  method = permutation.test.gbm, 
  las = 1
  )
```